---
title: "Homework 1"
subtitle: "Research Methods, Spring 2025"
author: "Genevieve DeBell"
format:
  pdf:
    output-file: "debell-g-hwk2-1"
    output-ext:  "pdf"
    header-includes:
      - \usepackage{float}
      - \floatplacement{table}{H}
---



```{r}
#| include: false

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate, readr, readxl, hrbrthemes, fixest,
               scales, gganimate, gapminder, gifski, png, tufte, plotly, OECD,
               ggrepel, survey, foreign, devtools, pdftools, kableExtra, modelsummary,
               kableExtra)
```


```{r}
#| include: false
#| eval: true
 
load("/Users/genevievedebell/Documents/GitHub/hwk2/submission3/results/Hwk2_workspace.Rdata")
load("/Users/genevievedebell/Documents/GitHub/hwk2/submission3/results/Hwk2-2_workspace.Rdata")
```

# 1. How many hospitals filed more than one report in the same year? Show your answer as a line graph of the number of hospitals over time.

```{r}
#| echo: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, ggplot2, dplyr, lubridate)
multi_report_counts <- duplicate.hcris %>%
group_by(fyear) %>%
summarise(num_hospitals = n_distinct(provider_number))

## Plot the number of hospitals filing multiple reports by year
ggplot(multi_report_counts, aes(x = fyear, y = num_hospitals)) +
geom_line(color = "blue", size = 1) +
geom_point(color = "red", size = 2) +
labs(
title = "Hospitals Filing Multiple Reports Per Year",
x = "Year",
y = "Number of Hospitals",
caption = "Source: HCRIS Data (1996 & 2010 Versions)"
) +
theme_minimal()
```

# 2. After removing/combining multiple reports, how many unique hospital IDs (Medicare provider numbers) exist in the data?
```{r}
#| echo: false
fig.unique <- final.hcris.data %>% group_by(year) %>%
  summarize(unique_hospital_count=n()) %>%
  ggplot(aes(x=as.factor(year), y=unique_hospital_count, group=1)) +
  geom_line() +
  labs(
    x="Year",
    y="Number of Hospitals",
    title=""
  ) + theme_bw() +
  scale_y_continuous(labels=scales::comma,limits=c(0,6500)) +
  theme(axis.text.x = element_text(angle=70, hjust=1))
print(fig.unique)
```

# 3. What is the distribution of total charges (tot_charges in the data) in each year?
```{r}
#| echo: false
charge.data <- final.hcris.data %>% 
  group_by(year) %>% 
  mutate(tot_charges_low=quantile(tot_charges, probs=0.01, na.rm=TRUE), 
         tot_charges_high=quantile(tot_charges, probs=0.99, na.rm=TRUE)) %>%
  filter(tot_charges<tot_charges_high, tot_charges>tot_charges_low, 
        !is.na(tot_charges), year>1997) %>% 
  mutate(log_charge=log(tot_charges))

ggplot(charge.data, aes(x = factor(year), y = tot_charges)) +
geom_violin(fill = "lightblue", color = "darkblue", alpha = 0.6) +
scale_y_log10() + # Apply log scale to handle skewed distributions
labs(
title = "Distribution of Total Charges by Year",
x = "Year",
y = "Total Charges (log scale)",
caption = "Source: HCRIS Data (1996 & 2010 Versions)"
) +
theme_minimal()
```

# 4. What is the distribution of estimated prices in each year?
```{r}
#| echo: false
price.data <- final.hcris.data %>%
  mutate(discount_factor=1 - tot_discounts / tot_charges,
    price_num = (ip_charges + icu_charges + ancillary_charges)*discount_factor - tot_mcare_payment,
    price_denom=tot_discharges - mcare_discharges,
    price=price_num/price_denom) %>% 
  filter(price_denom>100, !is.na(price_denom), 
    price_num>0, !is.na(price_num), 
    price<100000, 
    beds>30, !is.na(beds))

## Create a violin plot to show the distribution of estimated prices by year
ggplot(price.data, aes(x = as.factor(year), y = price)) +
  geom_violin(trim = TRUE, fill = "skyblue", color = "black") +
  labs(
    title = "Distribution of Estimated Prices by Year",
    x = "Year",
    y = "Price"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  ```

# 5. Calculate the average price among penalized versus non-penalized hospitals, only including observations from 2012. 
```{r}
#| echo: false
price.data <- price.data %>%
  mutate( discount_factor = 1-tot_discounts/tot_charges,
          price_num = (ip_charges + icu_charges + ancillary_charges)*discount_factor - tot_mcare_payment,
          price_denom = tot_discharges - mcare_discharges,
          price = price_num/price_denom)

final.hcris.2012 <- price.data %>% ungroup() %>%
  filter(price_denom>100, !is.na(price_denom), 
         price_num>0, !is.na(price_num),
         price<100000, 
         beds>30, year==2012) %>%  #<<
  mutate( hvbp_payment = ifelse(is.na(hvbp_payment),0,hvbp_payment),
          hrrp_payment = ifelse(is.na(hrrp_payment),0,abs(hrrp_payment)), #<<
    penalty = (hvbp_payment-hrrp_payment<0)) #<<

mean.pen <- round(mean(final.hcris.2012$price[which(final.hcris.2012$penalty==1)]),2)
mean.nopen <- round(mean(final.hcris.2012$price[which(final.hcris.2012$penalty==0)]),2)
cat("Average price among non-penalized hospitals:", mean.nopen, "\n")
cat("Average price among penalized hospitals:", mean.pen, "\n")
```

# 6. Split hospitals into quartiles based on bed size. To do this, create 4 new indicator variables, where each variable is set to 1 if the hospital’s bed size falls into the relevant quartile. Provide a table of the average price among treated/control groups for each quartile.
```{r}
#| echo: false
final.hcris.2012 <- final.hcris.2012 %>%
mutate(
hvbp_payment = ifelse(is.na(hvbp_payment), 0, hvbp_payment),
hrrp_payment = ifelse(is.na(hrrp_payment), 0, hrrp_payment),
penalty = (hvbp_payment + hrrp_payment) < 0
)

## Calculate bed size quartiles
bed_quartiles <- quantile(final.hcris.2012$beds, probs = c(0.25, 0.50, 0.75), na.rm = TRUE)

## Clean by bed size and price 
bed_lower <- quantile(final.hcris.2012$beds, 0.05, na.rm = TRUE)
bed_upper <- quantile(final.hcris.2012$beds, 0.95, na.rm = TRUE) 
price_cutoffs <- quantile(final.hcris.2012$price, probs = c(0.05, 0.95), na.rm = TRUE)

final.hcris.2012 <- final.hcris.2012 %>%
filter(beds > bed_lower & beds < bed_upper) %>%
filter(price >= price_cutoffs[1] & price <= price_cutoffs[2]) %>%
mutate(
Q1 = ifelse(beds <= bed_quartiles[1] & beds > 0, 1, 0),
Q2 = ifelse(beds > bed_quartiles[1] & beds <= bed_quartiles[2], 1, 0),
Q3 = ifelse(beds > bed_quartiles[2] & beds <= bed_quartiles[3], 1, 0),
Q4 = ifelse(beds > bed_quartiles[3], 1, 0)
) 

## Calculate average prices by quartile and penalty status
quartile_summary <- final.hcris.2012 %>%
mutate(bed_quartile = case_when(
Q1 == 1 ~ "Q1",
Q2 == 1 ~ "Q2",
Q3 == 1 ~ "Q3",
Q4 == 1 ~ "Q4"
)) %>%
group_by(bed_quartile, penalty) %>%
summarise(avg_price = mean(price, na.rm = TRUE), .groups = "drop") %>%
pivot_wider(names_from = penalty, values_from = avg_price, names_prefix = "penalty_")
quartile_summary <- quartile_summary %>%
  rename(
    `No Penalty` = `penalty_FALSE`, 
    `Penalty` = `penalty_TRUE`)
print(quartile_summary)
```

# 7. Find the average treatment effect using each of the following estimators, and present your results in a single table: Nearest neighbor matching (1-to-1) with inverse variance distance based on quartiles of bed size, nearest neighbor matching (1-to-1) with Mahalanobis distance based on quartiles of bed size, inverse propensity weighting, where the propensity scores are based on quartiles of bed size, simple linear regression, adjusting for quartiles of bed size using dummy variables and appropriate interactions as discussed in class.
```{r}
#| echo: false
install.packages("Matching")
install.packages("MatchIt")
install.packages("WeightIt")
install.packages("cobalt")
library(Matching)
library(cobalt)
library(dplyr)

final.hcris.2012 <- final.hcris.2012 %>%
  mutate(penalty = as.numeric(penalty))

## Nearest neighbor matching with inverse vairance distance based on quartiles of bed size

near.match <- Matching::Match(Y=final.hcris.2012$price,
                            Tr=final.hcris.2012$penalty,
                            X=(final.hcris.2012 %>% dplyr::select(Q1, Q2, Q3)),
                            M=1, 
                            Weight=1,
                            estimand="ATE",
                            ties=FALSE)
summary(near.match)

## Nearest Neighbor Matching (Mahalanobis)
mal.match <- Matching::Match(Y=final.hcris.2012$price,
                            Tr=final.hcris.2012$penalty,
                            X=(final.hcris.2012 %>% dplyr::select(Q1, Q2, Q3)),
                            M=1, 
                            Weight=2,
                            estimand="ATE",
                            ties=FALSE)
summary(mal.match)

### Calculate inverse propensity weights (IPW)
final.hcris.2012 <- final.hcris.2012 %>%
  mutate(ipw = case_when(
    penalty == 1 ~ 1 / ps,         # For treated group (penalty == 1)
    penalty == 0 ~ 1 / (1 - ps),   # For control group (penalty == 0)
    TRUE ~ NA_real_               # Handle any missing values
  ))

### Compute weighted average prices for treated (penalty == 1) and control (penalty == 0) groups
mean.t1 <- final.hcris.2012 %>% filter(penalty == 1) %>%
  dplyr::select(price, ipw) %>%
  summarize(mean_p = weighted.mean(price, w = ipw))

mean.t0 <- final.hcris.2012 %>% filter(penalty == 0) %>%
  dplyr::select(price, ipw) %>%
  summarize(mean_p = weighted.mean(price, w = ipw))

ipw.diff <- mean.t1$mean_p - mean.t0$mean_p
ipw.diff

## Simple Linear Regression
reg.data <- final.hcris.2012 %>% ungroup() %>%
  mutate(Q1_diff = penalty * (Q1 - mean(Q1)),
         Q2_diff = penalty * (Q2 - mean(Q2)),
         Q3_diff = penalty * (Q3 - mean(Q3)))

### Fit the regression model with quartile differences and penalty
reg <- lm(price ~ penalty + Q1 + Q2 + Q3 + 
            Q1_diff + Q2_diff + Q3_diff,
          data = reg.data)
summary(reg)

ate <- coef(reg)["penalty"]
ate

# Extract ATE estimates
ate_near_match <- near.match$est
ate_mal_match <- mal.match$est
ate_ipw_diff <- ipw.diff
ate_regression <- ate

# Create a data frame summarizing the results
ate_estimates <- data.frame(
  Method = c(
    "Nearest Matching (Inverse Variance)",
    "Nearest Matching (Mahalanobis Distance)",
    "Inverse Propensity Weighting (IPW)",
    "Linear Regression"
  ),
  ATE_Estimate = c(ate_near_match, ate_mal_match, ate_ipw_diff, ate_regression)
)

# Print the table using knitr::kable()
knitr::kable(ate_estimates, 
             caption = "ATE Estimates from Different Methods", 
             col.names = c("Method", "ATE Estimate"))
```

